Combining neural networks and geostatistics for landslide hazard assessment of San Salvador Metropolitan Area, El Salvador \newline \newline Combinando redes neuronales y geoestad\'{i}stica para evaluaci\'{o}n de deslizamientos de tierra de el \'{a}rea Metropolitana de San Salvador, El Salvador

### 2015-06-04 paper.tex
04 June 2015 10:25

\section{Introduction}
\label{sec:intr}
El Salvador, one of the smallest and most crowded nations in Central America, extends in Pacific coast about 240 kilometers westward from the Gulf of Fonseca to the border with Guatemala see Fig. \ref{fig:mass01}. El Salvador borders an active subduction boundary between Cocos and Caribbean plates located 30 km offshore. Therefore El Salvador is affected with high seismicity and volcanic activity related to this active boundary. There is two main sources of seismic activity, the upper interface thrust that coincides with the location of the recent volcanoes and the Benioff-Wadaty zones of subducted Cocos plate where deeper intraplate earthquakes occurs (\cite{dewey}). Due this tectonic and volcanic activity El Salvador has rugged relief. The main ranges cross the country with a rough west-east trend, parallel to the coast, these are separated from each other by faults and grabens. These ranges present several highly active volcanoes. The surface geology in El Salvador is almost entirely volcanic, dominated by upper Tertiary to Holocene volcanic rocks, only sparse outcrops of sedimentary and plutonic crops are located in the northern ranges, in the border with Honduras (\cite{weber}). Some of the most recent volcanic layers are formed by poor consolidated ashes and tuffs highly erodible (\cite{bommer}).

@@

@@

\section{Acknowledgment}   
The authors are indebted to M.SC. Giovanni Molina from MARN who provided the data and useful information about landslide modelling techniques.

@@

\bibitem{weber}
Weber H.S., Wiesemann G., Lorenz  W., Scmith-Thorn\'{e} M.
\newblock Mapa geol\'{o}gico de Republica de El Salvador
\newblock escala 1 : 100 000 (6 sheets).
\newblock \emph{ Bundesanstalt f\"ur Geowissenschaften und Rohstoffe, Hannover, Germany. }

@@

>>>>

\section{Introduction}
\label{sec:intr}
El Salvador, one of the smallest and most crowded nations in Central America, extends in Pacific coast about 240 kilometers westward from the Gulf of Fonseca to the border with Guatemala (see Fig \ref{fig:mass01}). El Salvador borders an active subduction boundary between Cocos and Caribbean plates located 30 km offshore. Therefore El Salvador is affected with high seismicity and volcanic activity related to this active boundary. There is two main sources of seismic activity, the upper interface thrust that coincides with the location of the recent volcanoes and the Benioff-Wadaty zones of subducted Cocos plate where deeper intraplate earthquakes occurs (\cite{dewey}). Due this tectonic and volcanic activity El Salvador has rugged relief. The main ranges cross the country with a rough west-east trend, parallel to the coast, these are separated from each other by faults and grabens. These ranges present several highly active volcanoes. The surface geology in El Salvador is almost entirely volcanic, dominated by upper Tertiary to Holocene volcanic rocks, only sparse outcrops of sedimentary and plutonic crops are located in the northern ranges, in the border with Honduras (\cite{weber}). Some of the most recent volcanic layers are formed by poor consolidated ashes and tuffs highly erodible (\cite{bommer}).

@@

\section{Study area}
\label{sec:studyarea}
The Metropolitan Area of San Salvador (MASS), is formed by 14 municipalities among them there is San Salvador, the capital of the country. It has 587 km2 and according the most recent census \cite{minecon} it has 1,566,629 inhabitans with a density of population of 2669 hab/km2.  According geography MASS is extended over a flat erosional surface (650-760 m a.s.l.), gently sloping to the east towards Ilopango Caldera also known as Ilopango Lake. MASS is bordered to the south by the Balsamo Range and San Jacinto Hill, to the west by San Salvador volcano and with Mariona Hills (refer to Fig \ref{fig:mass01}). Human settlement is not only restricted to the plain, but spreads up to surrounding heights and volcano flanks.

All rocks exposed in MASS have volcanic origin and consists of intercaled primary and reworked deposits from Late Tertiary to Holocene age \cite{schmidt1975}. The younger ones are known as Tierra Blanca pyroclastic ash deposits and forms a relative continuous layer in majority of MASS with an average thickness of 4 m \cite{schmidt1975}. These are deposits derived mainly from Ilopango caldera eruptions where they present more than 50 m of thickness \cite{schmidt1975}. Tierra Blanca soils is highly susceptible to erosion, one of the most common dangerous effects of the erosion in this area are landslides during heavy rainfall and earthquake ground shaking \cite{schmidt1975}. Chavez et al. 2014a identifies high and moderate erosion hazard in majority of MASS area. The more recent layer Tierra Blanca, known as Tierra Blanca Joven (TBJ) (Hernández 2004) present very poor geotechnical properties, especially when there is an increment of soil moisture (i.e. during rainy seasons), which results in high instability susceptibility (Chavez et al. 2014b and Rolo et al. 2004). Several incised rivers and ravines cross MASS area, due the large erosion rates of the area, natural slopes of these watercurses are often close to vertical and can reach highs of more than 10 m. Taking into account this property of volcanic soils, usually vertical slopes are cut to urban and road construction. As Bommer \& Rodriguez (2002) identify in Central America, although such slopes may remain stable for years, they may become unstable, abruptly and totally under the action of heavy rainfall or seismic shaking. Hernández (2004) carry out a detailed description of this process in TBJ, where these instabilities are very common. 
In study area fault trends are characterized by steep dip angles (650-900). The main fault system is characterized by an east-west trend; this trend is responsible of the steep northern slope of the El Balsamo Range.  The secondary systems are with north, northwest and northeast trends offset the main fault system and are more active in the present tectonic setting \cite{schmidt1975}. However, all the mentioned fault families appear active. Thus the several fault systems have developed at different times, but been repeatedly activated (Rymer 1987).   Finally there are also several ringlike faults that can be related to subsurface collapse due to volcanic tectonic subsidence \cite{schmidt1975}.
Last eruption of San Salvador volcano was 1917.  The majority of the magma's total volume (97\%) extruded from several vents in the Northwest flank of the volcano, outside of MASS. However a fraction of MASS is located under the proximal volcanic hazard zone according Sofield D. (2004). In the steep slopes  San Salvador volcano, especially of Picacho, the highest peak of the volcano, located in the eastern side present a high lahar hazard that threats MASS (Major et al. 2004). On the other hand one of the most destructive eruption in in the world was 430 A.D. TBJ eruption produced by the Ilopango Caldera. According Dull (2004), that eruption produced total devastation in whole MASS area.

@@

\section{Acknowledgment}   
The authors are indebted with MARN who provided the data and GIS software such as ILWIS and ARCGIS.

@@

\bibitem{weber}
Weber H.S., Wiesemann G., Lorenz  W., Scmith-Thom\'{e} M.
\newblock Mapa geol\'{o}gico de Republica de El Salvador
\newblock escala 1 : 100 000 (6 sheets).
\newblock \emph{ Bundesanstalt f\"ur Geowissenschaften und Rohstoffe, Hannover, Germany. }

\bibitem{weber}

\newblock Mapa geol\'{o}gico de Republica de El Salvador
\newblock escala 1 : 100 000 (6 sheets).
\newblock \emph{ Bundesanstalt f\"ur Geowissenschaften und Rohstoffe, Hannover, Germany. }

@@

%%%%%%%%%%%%%%  Study Area %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibitem{minecon}
Ministerio de Economia- Direcci\'{o}n General de Estad\'{i}stica y Censos
\newblock VI Censo de Población y V de Vivienda
\newblock \emph{2007-2008}.
\penalty0 Available from http://goo.gl/yvjzPy

\bibitem{schmidt1975}
Schmidt-Thom\'{e} M.
\newblock The geology in San Salvador area (El Salvador, Central America), a basis for city development and planning
\newblock \emph{Geol. Jb.}.
\penalty0 13, 207-228, 1975.

%Rymer M.J. (1987). The San Salvador Earthquake of October 10, 1986 – Geologic Aspects Earthquake Spectra, %3(3), 435-463.

%Sofield D. (2004) Eruptive history and volcanic hazards of Volcan San Salvador. Chapter 11 in Geological Society of America Special paper 375 in Natural Hazards in El Salvador of Rose W.I., Bommer J.J., López D.L., Carr M.J. and Major J.J. 147-158.
%
%
%
%
%
%Chávez J.A., Sebesta J., Kopecky L., López R. (2014a). Application of Geomorphologic Knowledge for Erosion Hazard Mapping. Natural Hazards. 71:1323–1354. DOI 10.1007/s11069-013-0948-8
%
%Chávez J.A.,  Landaverde J.M., Ayala O.E., Mendoza L.E. (2014b). Application of Constitutive Models in the Volcanic Tephra “Tierra Blanca Joven”. Ingeniería 24 (2).53-78. ISSN: 2215-2652.
%
%Dull R.A. (2004) Lessons from the mud, lessons from the Maya: Paleoecological records of the Tierra Blanca Joven eruption. Chapter 7 in Geological Society of America Special paper 375 in Natural Hazards in El Salvador of Rose W.I., Bommer J.J., López D.L., Carr M.J. and Major J.J. 237-244.
%
%Hernández W.E. (2004). Características Geomecánicas y Vulcanologicas de las Tefras Tierra Blanca Joven, Caldera de Ilopango, El Salvador. MSc Thesis. Universidad Politécnica de Madrid – Universidad Politécnica de El Salvador [cited 2015 May 31]. Available from http://portafolio.snet.gob.sv/digitalizacion/pdf/spa/doc00079/doc00079.htm
%
%Major J.J. Chilling S.P., Pullinguer C.R., Escobar C.D. (2004). Debris-flow hazards at San Salvador, San Vicente, and San Miguel volcanoes, El Salvador. Chapter 7 in Geological Society of America Special paper 375 in Natural Hazards in El Salvador of Rose W.I., Bommer J.J., López D.L., Carr M.J. and Major J.J. 89-108.
%
%Ministerio de Economia- Dirección General de Estadística y Censos (2008) VI Censo de Población y V de Vivienda . [cited 2015 May 31]. Available from http://www.censos.gob.sv/util/datos/Resultados\%20VI%20Censo%20de%20Poblaci%C3%B3n%20V%20de%20Vivienda%202007.pdfthe 
%Rymer M.J. (1987). The San Salvador Earthquake of October 10, 1986 – Geologic Aspects Earthquake Spectra, 3(3), 435-463.
%
%Rolo R., Bommer J.J., Houghton B.F., Vallance J.W., Berdousis P., Mavrommati C., Morphy W. (2004). Geologic and engineering characterization of Tierra Blanca pyroclastic ash deposits. Chapter 5 in Geological Society of America Special paper 375 in Natural Hazards in El Salvador of Rose W.I., Bommer J.J., López D.L., Carr M.J. and Major J.J. 55-67.
%
%Schmidt-Thomé M. (1975). The geology in San Salvador area (El Salvador, Central America), a basis for city development and planning. Geol. Jb., 13, 207-228.
%Sofield D. (2004) Eruptive history and volcanic hazards of Volcan San Salvador. Chapter 11 in Geological Society of America Special paper 375 in Natural Hazards in El Salvador of Rose W.I., Bommer J.J., López D.L., Carr M.J. and Major J.J. 147-158.
%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
